\chapter{Business Software}

\section{Introduction}

This lecture will go over three primary types of business productivity software most businesses use on a daily basis: 
\begin{enumerate}
	\item Word processing software: for typing and formatting text and images on a page (such as essays and reports) 
	\item Presentation software: for creating slides with information and graphics (such as for meetings and briefings)
	\item Spreadsheet software: for storing data, doing computations, and making graphs
\end{enumerate}

\section{Word Processing}

Word processing software is used for creating, editing, and formatting documents. Many tech companies have their own version of word processing software. A few more familiar examples include
\begin{itemize}
	\item Microsoft Word
	\item Apple Pages
	\item Google Docs
\end{itemize}

Beyond providing an interface to type the words of your report or essay into, word processors allow a variety of styling and formatting options to customize the document to look the way you like it. For example, all of the word processors above support
\begin{itemize}
	\item Automatic page numbering
	\item Adjustable margins
	\item Varieties of text size, text font, text color
	\item Addition of images and captions, along with image formatting
	\item Addition of tables and lists 
	\item Spelling and grammar checking
\end{itemize}

% Pop in  a picture

\section{Spreadsheets}

Spreadhseet software is used for dealing with data. A few examples of spreadsheet software include
\begin{itemize}
	\item Microsoft Excel
	\item Apple Numbers
	\item Google Sheets
\end{itemize}

Beyond providing a grid in which you can type numbers, spreadsheet software facilitate a variety of operations on the data. styling. A few key features supported by most spreadsheet applications include
\begin{itemize}
	\item Statistical analysis
	\item Sorting
	\item Creating charts and graphs
	\item Formatting (e.g. adding color, borders, etc.)
\end{itemize}

% maybe some activities here about what you would do with given data? 
% probably want to add pictures to illustrate each of the operations above 

\section{Presentations}

Presentation software is used for presenting data and graphics to an audience. Some common examples include
\begin{itemize}
	\item Microsoft Powerpoint
	\item Apple Keynote
	\item Google Slides
\end{itemize}

All of these software give the user a series of "slides" that can be clicked through during a presentation. It allows the presenter to focus the audience's attention on a single point at a time. The slides are not meant to contain all the information that the presenter wants to get across. Instead, the slides only mention the key takeaways, and the presenter should verbally fill in the details. For that reason, slides often consist of just a few bullet points and an image. Of course, there are a variety of creative ways to present - this is just the most common format!

A few particularly important features offered by almost all presentation software include
\begin{itemize}
	\item Automatic styling (color schemes, layouts, etc) 
	\item Add images and videos
	\item Slide transitions 
	\item Animations (e.g. showing one line of text on a slide at a time) 
\end{itemize}

\section{Project} 

